### 2011-02-11 htdocs/porous.tex
Updates to tex documents
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1271 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sections.tex
%
%  Content for Ludwig documentation.
%
%  $Id: sections.tex,v 1.14 2010-04-30 11:00:50 kevin Exp $
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2008 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\subsection{Note to Readers}

This document is intended as a starting point for a full-blown
specification of the next version of the Ludwig code, and everything
to do with it. It is (manifestly) not complete.

\subsection{Some History}

Early Lattice Gas and Lattice Boltzmann. Early Lattice Boltzmann at Edinburgh.
Ludwig.

\subsection{Aims}

We aim to provide a robust and portable code, written in C, which
can be used to perform serial and scalable parallel simulations of
complex fluid systems based around the Lattice Boltzmann method.

\subsection{Availability}

We now have a NeSCForge project account available at

\texttt{http://forge.nesc.ac.uk/projects/ludwig/}

This provides controlled access to known
persons and provide CVS, bug tracking, and so on. It is
\textbf{essential} that people sign up to this, as we must
move to a single supported version.

The NeSCForge CVS is not available to the public.

\subsection{Download}

\section{User Input}

By default the executable will look for an input file at run time.
By default, this file is named \texttt{input} and should be in the
current working directory. The alternative is to specify the name
of the file explicitly on the command line:
\begin{verbatim}
$ ./Ludwig.exe input_file
\end{verbatim}

The contents of the input file are made up of \textit{key value}
pairs with control the run-time behaviour. Blank lines and lines
begining with \# are ignored as comments. The following describes
the effect of various keys. Most keys have a default value which
will be used by the code if the corresponding key is not present
(or commented out) in the input file.

\subsection{Basic run parameters}

\inputkey{N\_cycles}

The number of lattice Boltzmann time steps to execute. Default
value:

\texttt{N\_cycles 0}

\inputkey{t\_start}

By default the code will start from time step $t = 0$. If you 
wish to restart from a previously saved configuration, set
the appropriate value for \texttt{t\_start}. The code will
execute \texttt{N\_cycles} steps starting from this point.

\inputkey{input\_config}

The file name stub for input configuration required for restarts.
For example, to restart from the file config.out100 previously
produced at the end of a 100 step run use
\begin{verbatim}
t_start 100
input_config config.out100
\end{verbatim}

If you are using parallel I/O, the filename stub is required; the
file extensions are computed automatically.

\inputkey{size}

Controls the number of lattice points in the $(x, y, z)$ directions,
respectively. Default value:

\texttt{size 64\_64\_64}

\inputkey{grid}

In parallel, this sets the processor Cartesian decomposition, i.e.,
the number of processors in each dimension of the Cartesian communicator.
The default value may be implementation dependent as it is that returned
by \texttt{MPI\_Dims\_create()}. For
example:
\begin{verbatim}
size 64_64_64
grid 4_2_1
\end{verbatim}
gives a total (physical) system size of 64 lattice sites in each
vdirection, decomposed across 4 processors in the x-direction, 2 in
the y-direction and 1 (no decomposition) in the z-direction. The
total number of processors must therefore be 8. The local domain
size per processor is then 16x32x64.

\subsection{Fluid properties}

\inputkey{viscosity}

Sets the LB fluid shear viscosity $\eta$ (and related relaxation times).
Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
(i.e., realaxation time equal to unity).

\inputkey{viscosity\_bulk}

Sets the LB bulk viscosity $\zeta$ (and related relaxation times).
This can be use if you are concerned with incompressiblity
violations, in which case $\zeta >> \eta$ can be useful. The
default value is $\zeta = \eta$.

\inputkey{temperature}

The LB 'temperature' using fluctuating hydrodynamics. Safe values
are $0.0001 > kT > 0$.

\inputkey{rho0}

Sets the mean density of the fluid. The default value is unity, and
it is recommended that the value should not be changed.

\subsection{Binary fluid parameters}

The are currently three choices of free energy relating to
various systems.

\subsubsection{Symmetric binary fluid}

\inputkey{free\_energy symmetric}

The following applies for the binary fluid problem with
compositional order parameter $\phi$ and free energy
(excluding the term in the density $\rho$):
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
\end{equation}

This is described in some detail
by Kendon et al \cite{viv}. The first two terms represent the bulk
contribution, whereas the term in $\kappa$ penalisies curvature in
the interface.

\inputkey{A}

The parameter $A$. Note $A < 0$.

\inputkey{B}

The parameter $B$. Note $B = -A$ for common usage, although this is
not enforced.

\inputkey{K}

The parameter $\kappa$, which is positive.

\subsubsection{Brazovskii free energy}

This is similar to the symetric, but with one extra term which
requires description.
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
\end{equation}

\subsubsection{Surfactant model free energy}

\inputkey{free\_energy surfactant}

This is for use with a two order parameter model the first of
which is the composition, as for the symmetric free energy,
while the second $\psi$ represents surfactant concentration.
The free energy density is made up of a number of terms:

\begin{equation}
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
\end{equation}
is the standard symmetric term related to the composition as above;

\begin{equation}
D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
\end{equation}
represents the energy of surfactant in the bluk phase with surfactant
concentration $0 < \psi < 1$ with a single parameter $D$;
\begin{equation}
-{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
-{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
\end{equation}
where the term in $\epsilon$ represents the energy reduction by
adsorbing surfactant at the interface and the term in $\beta$
provides [TO BE ACCOUNTED FOR]; finally, there is an extra term
\begin{equation}
+{\textstyle \frac{1}{2}} W \psi \phi^2
\end{equation}
which is included to stabilise the interface \cite{theissengompper}.

The corresponding keys for the input file are:

\inputkey{D}

is the bulk surfactant contribution parameter

\inputkey{epsilon}

is the parameter $\epsilon$

\inputkey{W}

is the parameter $W$

\inputkey{beta}

is the value of the parameter $\beta$ (not implemented yet).

\inputkey{phi\_b}

the initial, uniform, background surfactant concentration $\psi_b$.

\subsection{Other Binary Fluid Parameters}

\inputkey{mobility}

Sets the compositional order parameter mobility $M$ (uniform in space)
which appears in the Cahn-Hilliard equation.

\inputkey{mobilty\_psi}

Sets the (uniform) mobility for the surfactant as appropriate for
the Cahn-Hilliard equation.

\inputkey{phi0}

The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
The default value is zero, i.e., a symmetric 50:50 mixture by volume.

\inputkey{phi\_initialisation}

Determines how the compositional order parameter is initialisated
at the start of the run. If set to \texttt{spinodal} the value
is set to $\phi_0$ as set above plus or minus a random noise, the
magnitude of which is set by the value of the \texttt{noise} key.
If set to \texttt{block} a one-dimensional profile is set up in
the $z$-direction representing two blocks of fluid with
$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
appropriate width. The $\phi = -1$ section is in the middle.

\inputkey{noise}

The magnitude of the initial fluctuations in $\phi$ used to
initiate spinodal decompsotion.

\subsection{Colloidal particles}

There are quite a few options here.

\inputkey{colloid\_init}

This determines how the code initialises particles (or not). The
default is to have no particles (value \texttt{no\_colloids}). There
are two other options at the moment.

\texttt{colloid\_init    fixed\_number\_monodisperse}

initialises a fixed number of particles with randomly selected
positions. Note that there is no gaurentee that the particles
do not overlap. The number of particles is determined by
the value of the key \texttt{colloid\_no}.

\texttt{colloid\_init   fixed\_volume\_fraction\_monodisperse}

initialises a given volume fraction of particles on a (roughly)
bcc lattice. The particles will not overlap. The volume fraction
is set using the key \texttt{colloid\_vf}.

\inputkey{colloid\_a0}

Sets the nominal radius of the particle on the lattice $a_0$. This is
used to construct the links for BBL. The smallest acceptable LB
particle is $a_0 = 1.25$ \cite{nl02}. There are 'magic' values
including 1.25, 2.3, 3.7, 4.7 which minimise discretisation effects.

\inputkey{colloid\_ah}

Sets the hydrodynamic radius of the particles $a_h$. For all
reasonable applications choose $a_h = a_0$.

\inputkey{colloid\_mc\_steps}

It is possible to have the code follow the initialisation of
particle positions by a number of Monte Carlo steps. This can
help to remove overlaps (fixed\_number\_monodisperse) or
melt the lattice (fixed\_volume\_fraction\_monodisperse).

\inputkey{colloid\_gravity}

Sets a body force on each paarticle. Useful, for example, for
sedimentation. For example,

\inputkey{cell\_list\_min}

This sets the minimum allowable width of the cell list used to
store the particles. This should be as small as possible for
efficiency purposes but must be large enough to capture all
possible interactions.

\texttt{colloid\_gravity  0.0\_0.0\_-0.0001}

gives a constant force in the negative $z$-direction on each particle.
The code automatically computes the compensating body force on fluid nodes
required to give no net change in the total momentum of the system at
each time step.

\subsection{Active particles}

Not in general use yet.

\subsection{Particle-particle interactions}

\subsection{Boundary conditions}

Solid planar boundary walls can be added at the extremities of the
system (i.e., the volume of fluid is extactly that specified by
the \texttt{size} keyword). The wall can be in one, two or all three
coordinate direcctions, i.e., a 'slab' geometry, and 'duct' geometry
or a 'box' geometry, respectively.

The boundary walls are switched on via the keyword value pair

\inputkey{boundary\_walls\_on [no | yes]}

If \texttt{yes} then the geometry is specified with the keyword
as one of

\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}

Note that these are the only valid selections. \texttt{1\_1\_0}
gives periodic boundaries in the $x-$ and $y-$directions, with
walls at either edge in the $z-$direction. \texttt{1\_0\_0}
gives periodic conditions in the $x-$ direction only, while
\texttt{0\_0\_0} is a fully enclosed box.

In the case of the slab geometry, two further values are used so
that shear may be imparted to the system:

\inputkey{boundary\_speed\_bottom}

\inputkey{boundary\_speed\_top}

These values set the $x-$component of the wall velocity at
$z = z_{min}$ and $z = z_{max}$, respectively. Like all
velocities, these values must respect the low Mach number
constraint $u_x < c_s$.

\subsection{Input/Output and Parallel I/O}

\inputkey{freq\_statistics}

sets the frequency for reporting general information from
the code. (In general, the function
 \texttt{is\_statistics\_step()} returns true at this frequency.

\inputkey{freq\_measure}

sets the frequency of recording of colloid/order parameter output.
(In general, the function \texttt{is\_measurement\_step()} returns
true at this frequency.)

\inputkey{freq\_config}

set frequency of full configuration (checkpoint) dumps suitable
for restart.

\inputkey{freq\_phi}

set frequency of order parameter output.

\inputkey{phi\_format}

either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.

\inputkey{freq\_vel}

set frequancy of velocity output.

\inputkey{vel\_format}

either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.

\inputkey{config\_at\_end}

enforces full configuration dump at the end of execution. The
default is 'yes'.

\subsection{Lees-Edwards planes}

\inputkey{N\_LE\_plane}

sets the total number of Lees-Edwards planes. Default is zero.
The placing is as followings. The number of planes $n$ must
divide the lattice size in the $x$-direction to give an integer
$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.

\inputkey{LE\_plane\_vel}

sets the velocity of each plane relative to the lattice. All planes
have the same, constant, velocity.

\inputkey{LE\_init\_profile}

if set to 1, the fluid velocity is initialise to reflect a steady
state shear flow appropriate for the number of planes at the
given velocity. If set to zero, the fluid is initialised with
zero velocity.

\subsection{Porous Media}

@@

\subsubsection{File format}

@@

\subsubsection{Permeability calculations}

@@

Note that in the case of porous media with narrow channels at the grid
scale, the wall velocity can be dependent on the viscosity of the fluid.
This is an artefact of the bounce-back on links and will result in a
viscosity-dependent permeability (see, e.g., \cite{lipanmiller}). To minimise
this effect, it is recommended to switch on the correction to the ghost
mode relaxation time proposed by Ginzburg and d'Humi\`eres \cite{ginzburg}
which corrects the viscosity-dependence of the apparent boundary
position. This is done via the input switches

\texttt{ghost\_modes on\\
ginzburg\_dhumieres on}

and should recover a viscosity-independent permeability measurement
\cite{lipanmiller,chunladd}.

\subsubsection{Example: circular capillaries}

@@

\subsubsection{Example: rectangular capillaries}

@@

\subsubsection{Matching lattice and real units in porous media}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  porous.tex
%
%  Some considerations for porous media.
%
%  $Id$
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2011 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Porous Media}

@@

\subsection{File format}

@@

\subsection{Permeability calculations}

@@

Note that in the case of porous media with narrow channels at the grid
scale, the wall velocity can be dependent on the viscosity of the fluid.
This is an artefact of the bounce-back on links and will result in a
viscosity-dependent permeability (see, e.g., \cite{lipanmiller}).
To minimise this effect, e.g.,  Ginzburg and d'Humi\`eres \cite{ginzburg}
corrects the viscosity-dependence of the apparent boundary position
using a three relaxation time scheme. This is currently under
investigation.

\subsection{Example: circular capillaries}

@@

\subsection{Example: rectangular capillaries}

@@

\subsection{Matching lattice and real units in porous media}
